How and Why to run a Hack Week

### 2017-08-15 paper.tex
Line break after every full stop

As data becomes cheaper to gather and store, research across a wide range of disciplines has become increasingly reliant on computational workflows involving a familiarity with aspects of statistical modeling, machine learning, scalable computation, and related skills.
At the same time, formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, third-party workshops.
A well-known example is the Software Carpentry and Data Carpentry workshops, whose interdisciplinary workshops in research computing skills have reached more than 16,000 participants since its inception in 1998\footnote{\url{https://f1000research.com/articles/3-62/v2}}.
At the same time, there has been a rise in the number of domain-specific courses focusing on statistics and computation within their field.; in the Astronomy community, one of the longest-running has been at the Penn State Center for Astrostatistics *Summer School in Statistics for Astronomers*, consisting of ''intensive week covering basic statistical inference, several fields of applied statistics, and the R computing environment``\footnote{\url{http://astrostatistics.psu.edu/su16/}}. In the geospatial community, some software providers are taking the lead on educating the community, for example through the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}.
Another model of extra-curricular meeting is less pedagogical and more project-focused, perhaps typified by the dotAstronomy meetings, a week-long conference series which focuses specifically on leveraging web technologies for novel applications in Astronomy\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have started in conjunction with conferences, such Hack Days at the annual American Astronomical Society meetings, where Astronomers gather to work intensively on collaborative projects, the Brainhack hackathons that take place in conjunction with meetings of the Organization for Human Brain Mapping and the Society for Neuroscience\footnote{\url{https://gigascience.biomedcentral.com/articles/10.1186/s13742-016-0121-x}}, and a hackathon associate with the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}.
In broad-brush, pedagogically-focused workshops and summer schools tend to follow a classic academic model where novices learn a skillset from experts, while project-focused workshops focus on people collaboratively exercising the skillset they already have.
A disadvantage of the summer school model is that it can tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the hackathon model is the common perception (whether accurate or not) that the week is designed for experts in technical tools, which may discourage others from attending.
In 2014, we started Astro Hack Week to try to fill the gaps between these models. The hack week model combines pedagogy (often focused on statistical and computational techniques) with room for collaborative hacks or creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career.

@@

The recent surge in popularity of these events has resulted in a broad spectrum of ways to define the hackathon. Following the typology of \citep{Drouhard2017}, we assert that core elements of all hackathons include opportunities for networking, the strengthening of social ties and the building of community connections, both within and across disciplines. Building on these core elements, there are various implementations of the hackathon concept with respect to the overall purpose, mode of participation, style of work environment and motivation \citep{Drouhard2017}. ''Catalytic`` hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge. ''Contributive`` hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project. Finally, ''Communal`` hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

@@

An example of such events are the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/](http://www.imagexd.org/}} meetings held at the UC Berkeley Institute for Data Science in the Summer 2016, and at the UW eScience Institute in the Spring of 2017. These three-day events also included a mix of tutorials, talks, and joint work on projects, but in contrast to the hack weeks described in the present paper, these events included participants from a variety of different research fields, with the common thread being that they all have an interest in the analysis of images, as part their research. Thus, audience and speakers included researchers from wide array of domains: astronomy, neuroscience and geosciences, but also material science, chemistry, biology, medicine, engineering, and others. In addition, participants included computer vision researchers, who develop algorithms that can be applied to answer scientific questions. Similarly, a series of events at UC Berkeley (Text XD) focused on the analysis of text data across different domains of research.

@@

Aside from time, space plays a particularly important role in facilitating a successful hack week. Universities in general provide a convenient venue with existing structures to facilitate hack weeks (e.g. access to scientific publications, institutional support including staff and funding). On the other hand, most spaces in universities are designed for lectures, which are diametrically opposed to unconference-style events. Thus, finding an appropriate physical space within a university may be a challenge. However, with the rise of active learning as a preferred teaching methods, traditional lecture spaces are transforming into more flexible spaces that are generally appropriate for hack weeks. As a general rule, the smaller a hack week's emphasis on a taught component, the more flexible the space has to be, with ample opportunity for re-configuration. At Astro Hack Week, we have found it beneficial to spend one day mid-week at a different location (e.g. a company headquarter) to engage participants and break the routine. It is worth noting that collaborative spaces for non-traditional workshops are becoming more prevalent in academia (some attached to universities, some not) and may provide support and infrastructure that a traditional university location may not (examples are the Lorentz Center in Leiden, the Aspen Center something or other, and **others**?).

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear. Individuals with IS thus experience a fear of being ''found out``, shamed and expelled from their environment \citep{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{http://astrohackweek.org/blog/ethnographic-notes.html} suggested that hack weeks are an environment prone to a particular kind of IS: participants might feel the need to be experts in multiple aspects of the activities pursued during a hack week: expertise in a scientific domain, as well as expertise in a the variety of technical tools used.  This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee. This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \citep{Parkman2016-ro}. Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations. Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques. With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS. More generally, being open about the presence and prevalence of IS can help participants feel more at ease (e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day.  I think it helped ease the feeling!!``}) . Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking.

@@

\subsection*{Box 2: Examples of Hack Week Outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used the opportunity of Astro Hack Week to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing well-tested, well-documented implementations of algorithms for time series analysis often used in X-ray astronomy.  
The start of this project was facilitated by the collaborative environment at Astro Hack Week, including expertise in how to start/run open-source projects, role models of successful projects, and an environment encouraging scientific risk taking.
Since its beginnings at Astro Hack Week, Stingray has matured into an enduring collaboration within the community with five active maintainers, a number of contributors and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week} 
In 2016, a Geo Hack Week project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States. The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine. George Richardson, one of the project leads, now works for a renewable resource company in Seattle. More information can be found [here](http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/).
\subsubsection*{Example 3: Neuro Hack Week} 
Motion of study participants inside of the MRI machine is a major concern in neuroimaging studies.
This is a particular concern in studies in which children or patients are studied, as they are more likely to move.
In the 2016 Neuro Hack Week one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off were varied.
They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four different researchers from four different institutions in two different countries) continued to work on this project remotely after the Neuro Hack Week was over, and they eventually wrote and published a paper describing these results in the open access journal Research Ideas and Outcomes \citep{leonard2017}.

@@

The fast-paced changes of the computational and methodological landscape require traditional fields of science to rapidly adapt to new data analysis challenges. Traditional modes of learning, including university curricula, are often too slow to incorporate new developments on short enough time scales to meet their acute need in scientific advancement. To address this imbalance, new types of workshops, including unconferences, hackathons and bootcamps, have been developed in recent years in various scientific disciplines to exist alongside with and support the existing structure of academic conferences. 
Here, we introduce one such concept, hack weeks, and detail the underlying philosophical ideas along with experiences from events held in three different fields.

@@

\begin{addendum}
 \item Anyone thanked here: *Feel free to move up to the Author list!* It is a pleasure to thank Laura Norén (NYU) for help on ethics and IRB. This work was partially supported by the Moore-Sloan Data Science Environments at UC Berkeley, New York University, and the University of Washington. Neurohackweek is supported through a grant from the National Institute for Mental Health (1R25MH112480). Daniela Huppenkothen is partially supported by the James Arthur Postdoctoral Fellowship at NYU.
 \item[Competing Interests] The authors declare that they have no
competing financial interests.
 \item[Correspondence] Correspondence and requests for materials
should be addressed to A.B.C.~(email: myaddress@nowhere.edu).
\end{addendum}

>>>>

As data becomes cheaper to gather and store, research across a wide range of disciplines has become increasingly reliant on computational workflows involving a familiarity with aspects of statistical modeling, machine learning, scalable computation, and related skills.
At the same time, formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, third-party workshops.
A well-known example is the Software Carpentry and Data Carpentry workshops, whose interdisciplinary workshops in research computing skills have reached more than 16,000 participants since its inception in 1998\footnote{\url{https://f1000research.com/articles/3-62/v2}}.
At the same time, there has been a rise in the number of domain-specific courses focusing on statistics and computation within their field. 
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, and more project-focused than pedagogical meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have started in conjunction with conferences, such Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons that take place in conjunction with meetings of the Organization for Human Brain Mapping and the Society for Neuroscience\footnote{\url{https://gigascience.biomedcentral.com/articles/10.1186/s13742-016-0121-x}}, and a hackathon associate with the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}.
In broad-brush, pedagogically-focused workshops and summer schools tend to follow a classic academic model where novices learn a skillset from experts, while project-focused workshops focus on people collaboratively exercising the skillset they already have.
A disadvantage of the summer school model is that it can tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the hackathon model is the common perception (whether accurate or not) that the week is designed for experts in technical tools, which may discourage others from attending.
In 2014, we started Astro Hack Week to try to fill the gaps between these models. 
The hack week model combines pedagogy (often focused on statistical and computational techniques) with room for collaborative hacks or creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career.

@@

%The recent surge in popularity of these events has resulted in a broad spectrum of ways to define the hackathon. 
Following the typology of \citep{Drouhard2017}, we assert that core elements of all hackathons include opportunities for networking, the strengthening of social ties and the building of community connections, both within and across disciplines. 
Building on these core elements, there are various implementations of the hackathon concept with respect to the overall purpose, mode of participation, style of work environment and motivation \citep{Drouhard2017}. 
''Catalytic`` hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge. 
''Contributive`` hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project. 
Finally, ''Communal`` hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

@@

An example of such events are the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/}} meetings held at the UC Berkeley Institute for Data Science in the Summer 2016, and at the UW eScience Institute in the Spring of 2017. 
These three-day events also included a mix of tutorials, talks, and joint work on projects, but in contrast to the hack weeks described in the present paper, these events included participants from a variety of different research fields, with the common thread being that they all have an interest in the analysis of images, as part their research. 
Thus, audience and speakers included researchers from wide array of domains: astronomy, neuroscience and geosciences, but also material science, chemistry, biology, medicine, engineering, and others. 
In addition, participants included computer vision researchers, who develop algorithms that can be applied to answer scientific questions. 
Similarly, a series of events at UC Berkeley (Text XD) focused on the analysis of text data across different domains of research.

@@

Aside from time, space plays a particularly important role in facilitating a successful hack week. 
Universities in general provide a convenient venue with existing structures to facilitate hack weeks (e.g. access to scientific publications, institutional support including staff and funding). 
On the other hand, most spaces in universities are designed for lectures, which are diametrically opposed to unconference-style events. 
Thus, finding an appropriate physical space within a university may be a challenge. 
However, with the rise of active learning as a preferred teaching methods, traditional lecture spaces are transforming into more flexible spaces that are generally appropriate for hack weeks. 
As a general rule, the smaller a hack week's emphasis on a taught component, the more flexible the space has to be, with ample opportunity for re-configuration. 
At Astro Hack Week, we have found it beneficial to spend one day mid-week at a different location (e.g. a company headquarter) to engage participants and break the routine. 
It is worth noting that collaborative spaces for non-traditional workshops are becoming more prevalent in academia (some attached to universities, some not) and may provide support and infrastructure that a traditional university location may not.

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear. 
Individuals with IS thus experience a fear of being ''found out``, shamed and expelled from their environment \citep{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{\url{http://astrohackweek.org/blog/ethnographic-notes.html}} suggested that hack weeks are an environment prone to a particular kind of IS: participants might feel the need to be experts in multiple aspects of the activities pursued during a hack week: expertise in a scientific domain, as well as expertise in a the variety of technical tools used. 
This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee. 
This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \citep{Parkman2016-ro}. 
Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations.
 Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques. 
With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS. 
More generally, being open about the presence and prevalence of IS can help participants feel more at ease (e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day. 
I think it helped ease the feeling!!``}). 
Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking.

@@

\subsection*{Box 2: Examples of Hack Week Outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used the opportunity of Astro Hack Week to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing well-tested, well-documented implementations of algorithms for time series analysis often used in X-ray astronomy.  
The start of this project was facilitated by the collaborative environment at Astro Hack Week, including expertise in how to start/run open-source projects, role models of successful projects, and an environment encouraging scientific risk taking.
Since its beginnings at Astro Hack Week, Stingray has matured into an enduring collaboration within the community with five active maintainers, a number of contributors and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week} 
In 2016, a Geo Hack Week project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States. 
The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine\footnote{\url{http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/}}. 
George Richardson, one of the project leads, now works for a renewable resource company in Seattle.
\subsubsection*{Example 3: Neuro Hack Week} 
Motion of study participants inside of the MRI machine is a major concern in neuroimaging studies.
This is a particular concern in studies in which children or patients are studied, as they are more likely to move.
In the 2016 Neuro Hack Week one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off were varied.
They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four different researchers from four different institutions in two different countries) continued to work on this project remotely after the Neuro Hack Week was over, and they eventually wrote and published a paper describing these results in the open access journal Research Ideas and Outcomes \citep{leonard2017}.

@@

The fast-paced changes of the computational and methodological landscape require traditional fields of science to rapidly adapt to new data analysis challenges. 
Traditional modes of learning, including university curricula, are often too slow to incorporate new developments on short enough time scales to meet their acute need in scientific advancement. 
To address this imbalance, new types of workshops, including unconferences, hackathons and bootcamps, have been developed in recent years in various scientific disciplines to exist alongside with and support the existing structure of academic conferences. 
Here, we introduce one such concept, hack weeks, and detail the underlying philosophical ideas along with experiences from events held in three different fields

@@

\begin{addendum}
 \item{Anyone thanked here: *Feel free to move up to the Author list!* It is a pleasure to thank Laura Norén (NYU) for help on ethics and IRB. This work was partially supported by the Moore-Sloan Data Science Environments at UC Berkeley, New York University, and the University of Washington. Neurohackweek is supported through a grant from the National Institute for Mental Health (1R25MH112480). Daniela Huppenkothen is partially supported by the James Arthur Postdoctoral Fellowship at NYU.}
 \item[Competing Interests]{The authors declare that they have no competing financial interests.}
 \item[Correspondence]{Correspondence and requests for materials should be addressed to D.Huppenkothen.~(email: daniela.huppenkothen@nyu.edu).}
\end{addendum}
